---
format: md
editor: visual
---

# Calculate daily means from TOMST data

Written by: Katrín Björnsdóttir (katrin.bjornsdottir\@bioenv.gu.se), 20. March 2025

This tutorial demonstrates how to calculate daily and monthly means from raw TOMST logger readings.

## Prepare the data

1.  Load packages *(make sure you have all packages installed already)*

```{r}
library(dplyr)
library(data.table)
library(lubridate)
library(purrr)
library(tidyverse)
```

2.  Load in the cleaned raw TOMST readings.

```{r, message = FALSE}
tomst_data <- read_csv("Outputs/Data/output_tomst_test.csv") %>%
  mutate(datetime = with_tz(datetime, tzone = "Europe/Stockholm"))
  
```

## Calculate daily means (and other useful variables)

```{r, message = FALSE, warning = FALSE}
daily_means <- tomst_data %>%
  mutate(date = as_date(datetime)) %>%
  group_by(plot_id, date) %>%
  mutate(
    roll_diff_T1 = T1 - lead(T1), # calculates the difference in soil temp between days
    roll_diff_T2 = T2 - lead(T2), # calculates the difference in surf temp between days
    roll_diff_T3 = T3 - lead(T3)  # calculates the difference in air temp between days
    ) %>%
  summarise(
    soil_mean = mean(T1), # daily mean temperature
    soil_median = median(T1), # daily median temperature
    soil_sd = sd(T1), # daily standard deviation
    soil_min = min(T1), # minimum temperature per day
    soil_max = max(T1), # maximum temperature per day
    soil_quantile_95 = quantile(T1, probs = 0.95), # 95% upper quantile temperature
    soil_quantile_5 = quantile(T1, probs = 0.05),  # 5% lower quantile temperature
    surf_mean = mean(T2),
    surf_median = median(T2),
    surf_sd = sd(T2),
    surf_min = min(T2),
    surf_max = max(T2),
    surf_quantile_95 = quantile(T2, probs = 0.95),
    surf_quantile_5 = quantile(T2, probs = 0.05),
    air_mean = mean(T3),
    air_median = median(T3),
    air_sd = sd(T3),
    air_min = min(T3),
    air_max = max(T3),
    air_quantile_95 = quantile(T3, probs = 0.95),
    air_quantile_5 = quantile(T3, probs = 0.05),
    corr_soil_air = cor(T1, T3, use = "pairwise.complete.obs"), # daily correlation between soil and air temperature
    max_roll_diff_soil = max(roll_diff_T1, na.rm = TRUE), # maximum daily difference
    max_roll_diff_surf = max(roll_diff_T2, na.rm = TRUE), # maximum daily difference
    max_roll_diff_air = max(roll_diff_T3, na.rm = TRUE),  # maximum daily difference
    moist_mean = mean(moist),
    moist_median = median(moist),
    moist_sd = sd(moist),
    moist_min = min(moist),
    moist_max = max(moist),
    moist_quantile_95 = quantile(moist, probs = 0.95),
    moist_quantile_5 = quantile(moist, probs = 0.05)
    ) %>%
  mutate(
    soil_air_diff = air_mean - soil_mean,
    soil_air_max_diff = air_max - soil_max
    )
```

## Calculate monthly means (and other useful variables)

```{r}
monthly_means <- tomst_data %>%
  mutate(date = as_date(datetime),
         # Extract year and month
         year = year(date),
         month = month.name[month(date)]
  ) %>%
  group_by(plot_id, year, month) %>%
  summarise(
    # summarise how many days are included in the monthly mean summary to make sure the data across months are comparable
    nr_days = n_distinct(date),
    soil_mean = mean(T1), # monthly mean temperature
    soil_median = median(T1), # monthly median temperature
    soil_sd = sd(T1), # monthly standard deviation
    soil_min = min(T1), # monthly minimum temperature
    soil_max = max(T1), # monthly maximum temperature
    soil_quantile_95 = quantile(T1, probs = 0.95), # monthly 95% upper quantile
    soil_quantile_5 = quantile(T1, probs = 0.05), # monthly 5% lower quantile
    surf_mean = mean(T2),
    surf_median = median(T2),
    surf_sd = sd(T2),
    surf_min = min(T2),
    surf_max = max(T2),
    surf_quantile_95 = quantile(T2, probs = 0.95),
    surf_quantile_5 = quantile(T2, probs = 0.05),
    air_mean = mean(T3),
    air_median = median(T3),
    air_sd = sd(T3),
    air_min = min(T3),
    air_max = max(T3),
    air_quantile_95 = quantile(T3, probs = 0.95),
    air_quantile_5 = quantile(T3, probs = 0.05),
    moist_mean = mean(moist),
    moist_median = median(moist),
    moist_sd = sd(moist),
    moist_min = min(moist),
    moist_max = max(moist),
    moist_quantile_95 = quantile(moist, probs = 0.95),
    moist_quantile_5 = quantile(moist, probs = 0.05)
  )
```

## Save daily and monthly means outputs

```{r}
write_csv(daily_means, "Outputs/Data/output_daily_means.csv")

write_csv(monthly_means, "Outputs/Data/output_monthly_means.csv")
```
